Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-06-22 article/article.tex
Commit 2024-06-22 19:11

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Long term cloud enhancement events of global solar irradiance over Thessaloniki, Greece},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

@@

  \title{Long term cloud enhancement events of global solar irradiance
over Thessaloniki, Greece}
    \author[LAP]{Athanasios N. Natsis%
  \corref{cor1}%
  }
   \ead{natsisphysicist@gmail.com} 
    \author[LAP]{Alkiviadis Bais%
  %
  }
   \ead{abais@auth.gr} 
    \author[LAP]{Charikleia Meleti%
  %
  }
   \ead{meleti@auth.gr} 
      \affiliation[LAP]{
    organization={Laboratory of Atmospheric Physics, Physics Department,
Aristotle University of
Thessaloniki},city={Thessaloniki},postcode={54124},country={Greece},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

@@

In this study, we evaluate the effects CE on GHI by investigating their
frequency of occurrence, intensity, and duration in a thirty-year period
of GHI observations at Thessaloniki, Greece. We used modeled clear sky
irradiance as a baseline to identify cloud enhancements and we
determined long term trends of the above mentioned metrics, their
climatology and some general characteristics. To our knowledge there are
no other studies that provide trends from such long dataset.

@@

\hypertarget{long-term-change-of-clear-sky-irradiance}{%
\subsubsection{Long term change of clear sky
irradiance}\label{long-term-change-of-clear-sky-irradiance}}

The above clear sky reference values are based on the climatological AOD
and WC, hence they cannot describe accurately the long term variation of
\(\text{GHI}_\text{ref}\) due to changes in the two atmospheric
constituents, mainly AOD. As reported by \citet{Natsis2023}, there is a
long term brightening effect in the GHI data of Thessaloniki for the
period 1993 -- 2023, which for clear-sky data was attributed to
long-term changes in aerosol effects. Therefore, an adjustment of the
\(\text{GHI}_\text{ref}\) during the period of study was made using
simulation with the RTM based on the long-term variations of the AOD. As
AERONET data start only in 2003, we used for the period 1993 -- 2005
estimates of changes in AOD at \(340\,\text{nm}\) derived from a
collocated Brewer spectrophotometer \citep{Kazadzis2007} to calculate
the trend in \(\text{GHI}_\text{ref}\) during this period.

According to this study, in the period 1997 -- 2005 the mean AOD at
\(340\,\text{nm}\) is \(0.403\) with a trend of \(-3.8\pm0.93\,\%\) per
year, corresponding to a change of \(0.0153\) per year. Using an
Ångström coefficient \(\alpha = 1.6\), this translates to a change in
the Ångström coefficient \(\beta=0.00272\) per year (or \(\beta=0.084\)
in 1997 and \(\beta=0.059\) in 2005). Simulations with uvspec for the
above Ångström coefficients, with WC of \(15.7\,\text{mm}\) and
\(15.4\,\text{mm}\) for 1997 and 2005 respectively and for a SZA of
\(55^\circ\) reveal a trend of \(+0.21\,\%\) per year in
\(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was chosen as
representative of all days in the year in order to get a rough estimate
of the annually averaged change in clear sky irradiance. For the period
2005 -- 2023 we used the mean monthly values of AOD and WC from AERONET
in a similar simulation scheme to calculate the monthly mean of clear
sky irradiance, and finally the trend of \(+0.14\,\%\) per year. We
applied these two long term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), in order to create a more realistic
representation of the clear-sky irradiance for the whole period of
study.

@@

\caption{Simulated long term change in clear sky irradiance relative to the climatological values due to changes in AOD.}\label{fig:CS-change}
\end{figure}

@@

In this study our main focus was to quantify over irradiance events
related to CEs. A key issue for achieving this goal is to define a
threshold for the CE identification, representative of the clear-sky
irradiance at the time of each GHI measurement. This depends on the
selection of the appropriate atmospheric parameterization for the RTM
simulations. The implementation of the long term change of AOD,
discussed in section\nobreakspace{}\ref{rtmcs}, allows capturing a large
part of the natural variability of clear-sky GHI. However, the
short-term variability of AOD cannot be taken adequately into account
when using monthly values for the model simulations. We tried different
approaches to increase the robustness of the methodology and compensate
for the limited accuracy of the RTM input data and the unpredictable
natural variability of the atmosphere.

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factors
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
increased the \(\text{GHI}_\text{ref}\) by (\(4\,\%\)) with an
additional constant offset of \(15\,\text{W}/\text{m}^2\), as described
in Equation\nobreakspace\ref{eq:CE4}. \begin{equation}
\text{CE} : \text{E}_\text{i} > 15 + 1.04 \cdot \text{E}_\text{CSm,i} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation} where: \(\text{E}\) the measured irradiance,
\(\text{E}_\text{CSm}\) the selected modelled clear sky irradiance, and
\(i\) each of the one-minute observation. This is the criterion of our
CE identification.

These values were determined through the implementation of an empirical
method with manual inspection of the CE identification results on
selected days of the whole dataset. We used seven sets of days with
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed in some of the edge cases, we
also used images from a sky-camera to further aid the decision of the
manual inspection.

@@

The distribution of the number of CE events as a function of OIR is
shown in Figure\nobreakspace{}\ref{fig:ovir-distribution}. Apparently,
there is an inverse relation between the frequency of CE events and OIR
with an exponential-like decline. This is expected, as the stronger the
CE events become, the rarer are the conditions favoring the occurrence
of CE events. For the majority (over \(62\,\%\)) of the CE events the
OIR is below the long term average of \(39.9\,\text{W}/\text{m}^2\),
while about \(8.1\,\%\) of the events correspond to OIR larger than
\(100\,\text{W}/\text{m}^2\) and up to highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OIR
values, has been reported by \citet{Vamvakas2020} for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

An interpretation of the CE trends, shows that the interaction of GHI
with the clouds, through this 30 year period, is a dynamic phenomenon
that's needs further investigation. Although, to approach it, we need
more long term observations of AOD and clouds characteristics.

>>>>

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

@@

  \title{Long-term cloud enhancement events of global solar irradiance
over Thessaloniki, Greece}
    \author[LAP]{Athanasios N. Natsis%
  \corref{cor1}%
  }
   \ead{natsisphysicist@gmail.com} 
    \author[LAP]{Alkiviadis Bais%
  %
  }
   \ead{abais@auth.gr} 
    \author[LAP]{Charikleia Meleti%
  %
  }
   \ead{meleti@auth.gr} 
      \affiliation[LAP]{
    organization={Laboratory of Atmospheric Physics, Physics Department,
Aristotle University of
Thessaloniki},city={Thessaloniki},postcode={54124},country={Greece},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

@@

In this study, we evaluate the effects CE on GHI by investigating their
frequency of occurrence, intensity, and duration in a thirty-year period
of GHI observations at Thessaloniki, Greece. We used modeled clear sky
irradiance as a baseline to identify cloud enhancements and we
determined long-term trends of the above mentioned metrics, their
climatology and some general characteristics. To our knowledge there are
no other studies that provide trends from such long dataset.

@@

\hypertarget{long-term-change-of-clear-sky-irradiance}{%
\subsubsection{Long-term change of clear sky
irradiance}\label{long-term-change-of-clear-sky-irradiance}}

The above clear sky reference values are based on the climatological AOD
and WC, hence they cannot describe accurately the long-term variation of
\(\text{GHI}_\text{ref}\) due to changes in the two atmospheric
constituents, mainly AOD. As reported by \citet{Natsis2023}, there is a
long-term brightening effect in the GHI data of Thessaloniki for the
period 1993 -- 2023, which for clear-sky data was attributed to
long-term changes in aerosol effects. Therefore, an adjustment of the
\(\text{GHI}_\text{ref}\) during the period of study was made using
simulation with the RTM based on the long-term variations of the AOD. As
AERONET data start only in 2003, we used for the period 1993 -- 2005
estimates of changes in AOD at \(340\,\text{nm}\) derived from a
collocated Brewer spectrophotometer \citep{Kazadzis2007} to calculate
the trend in \(\text{GHI}_\text{ref}\) during this period.

According to this study, in the period 1997 -- 2005 the mean AOD at
\(340\,\text{nm}\) is \(0.403\) with a trend of \(-3.8\pm0.93\,\%\) per
year, corresponding to a change of \(0.0153\) per year. Using an
Ångström coefficient \(\alpha = 1.6\), this translates to a change in
the Ångström coefficient \(\beta=0.00272\) per year (or \(\beta=0.084\)
in 1997 and \(\beta=0.059\) in 2005). Simulations with uvspec for the
above Ångström coefficients, with WC of \(15.7\,\text{mm}\) and
\(15.4\,\text{mm}\) for 1997 and 2005 respectively, also taken from
AERONET, and for a SZA of \(55^\circ\) reveal a trend of \(+0.21\,\%\)
per year in \(\text{GHI}_\text{ref}\). The SZA of \(55^\circ\) was
chosen as representative of all days in the year in order to get a rough
estimate of the annually averaged change in clear sky irradiance. For
the period 2005 -- 2023 we used the mean monthly values of AOD and WC
from AERONET in a similar simulation scheme to calculate the monthly
mean of clear sky irradiance, and finally the trend of \(+0.14\,\%\) per
year. We applied these two long-term changes (see
Figure\nobreakspace{}\ref{fig:CS-change}) to the climatological
\(\text{GHI}_\text{ref}\), in order to create a more realistic
representation of the clear-sky irradiance for the whole period of
study.

@@

\caption{Simulated long-term change in clear sky irradiance relative to the climatological values due to changes in AOD.}\label{fig:CS-change}
\end{figure}

@@

In this study our main focus was to quantify over irradiance events
related to CEs. A key issue for achieving this goal is to define a
threshold for the CE identification, representative of the clear-sky
irradiance at the time of each GHI measurement. This depends on the
selection of the appropriate atmospheric parameterization for the RTM
simulations. The implementation of the long-term change of AOD,
discussed in section\nobreakspace{}\ref{rtmcs}, allows capturing a large
part of the natural variability of clear-sky GHI. However, the
short-term variability of AOD cannot be taken adequately into account
when using monthly values in the model simulations. We tried different
approaches in order to strengthen the robustness of the methodology and
compensate for the limited accuracy of the RTM input data and the
unpredictable natural variability of the atmosphere.

First, we evaluated the performance of the modelled
\(\text{GHI}_\text{ref}\) in relation to the measured GHI for different
levels of atmospheric clearness, by using in the RTM the monthly
climatological AOD and WC, less their respective standard deviations.
These values represent typical atmospheres in Thessaloniki with lower
than average load of aerosols and humidity, which are the main factors
that attenuate the GHI, excluding clouds. With this approach the
simulated \(\text{GHI}_\text{ref}\) should be generally greater than the
measured GHI when aerosols are more abundant. To compensate for this, we
defined the following threshold \(\text{E}\) to compare the measured
\(\text{GHI}\): \begin{equation}
\text{CE} : \text{E} > 15 + 1.04 \cdot \text{GHI}_\text{ref} \,\,[\text{W}/\text{m}^2] \label{eq:CE4}
\end{equation} This is the criterion of our CE identification.

These values were determined through the implementation of an empirical
method with manual inspection of the CE identification results on
selected days of the whole dataset. We used seven sets of days with
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed, we also used images from a
sky-camera to further aid the decision of the manual inspection.

@@

The distribution of the number of CE events as a function of OIR is
shown in Figure\nobreakspace{}\ref{fig:ovir-distribution}. Apparently,
there is an inverse relation between the frequency of CE events and OIR
with an exponential-like decline. This is expected, as the stronger the
CE events become, the rarer are the conditions favoring the occurrence
of CE events. For the majority (over \(62\,\%\)) of the CE events the
OIR is below the longnterm average of \(39.9\,\text{W}/\text{m}^2\),
while about \(8.1\,\%\) of the events correspond to OIR larger than
\(100\,\text{W}/\text{m}^2\) and up to highest value of
\(412.4\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OIR
values, has been reported by \citet{Vamvakas2020} for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

An interpretation of the CE trends, shows that the interaction of GHI
with the clouds, through this 30 year period, is a dynamic phenomenon
that's needs further investigation. Although, to approach it, we need
more long-term observations of AOD and clouds characteristics.
